R Notebook

### 2021-06-20 Exemples and Implementations.Rmd
nuovi files

```
\
\textbf{Particle Filter}
\
With reference to the random walk plus noise of section XX, let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \frac{1}{N}\sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. For $t=1,...,n$ where $n$ is the length of the sample, at any iteration
\begin{itemize}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\end{itemize}
In addition, when $ESS<ESS_{0}$ \footnote{In our example we fix $ESS_{0}=N/2$, this is an arbitrary common rule of thumb.}, resempling applies
\begin{itemize}
\item Draw a sample of size N, $x_{t}^{(1)},...,x_{t}^{(N)}$,from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{itemize}
```{r}
PFfun<-function(data,N,m0,C0,tau,sigma){
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

@@

    if(ESS<(N/2)){
      index<-sample(N,size=N,replace=T,prob=w)
      x<-x[index]
      w<-rep(1/N,N)
    }else{}

@@

```{r}
GPFfun<-function(data,N,m0,C0,tau,sigma){
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

@@

```{r}
APFfun<-function(data,N,m0,C0,tau,sigma){
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

@@

    if(ESS<(N/2)){
      index<-sample(N,size=N,replace=T,prob=w)
      x1<-x1[index]
      w<-rep(1/N,N)
    }else{}

@@

\end{itemize}
```{r}
LWfun<-function(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif){

  xs     = rnorm(N,m0,sqrt(C0))
  if(unif==T){
  pars   = cbind(runif(N,0,10),runif(N,0,10))}else{}
  pars   = cbind(rgamma(N,shape=alphav,scale=betav),rgamma(N,shape=alphaw,scale=betaw))
  a      = (3*delta-1)/(2*delta)
  h2     = 1-a^2
  parss  = array(0,c(N,2,n))
  xss    = NULL
  ws     = NULL
  ess    = NULL
  w      = rep(1/N,N)
  for (t in 1:length(data)){
    meanV = weighted.mean(pars[,1],w)
    varV  = weighted.mean((pars[,1]-meanV)^2,w)
    meanW = weighted.mean(pars[,2],w)
    varW  = weighted.mean((pars[,2]-meanW)^2,w)

@@

    if(ESS<(N/2)){
      index<-sample(N,size=N,replace=T,prob=w)
      xs<-xs[index]
      pars<-pars[index,]
      w<-rep(1/N,N)
    }else{
      xs<-xs
      pars<-pars
    }

@@

>>>>

```
\
\textbf{Particle Filter}
\
With reference to the random walk plus noise of section XX, let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \frac{1}{N}\sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. For $t=1,...,n$ where $n$ is the length of the sample, at any iteration
\begin{itemize}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\end{itemize}
In addition, when $ESS<ESS_{0}$ \footnote{In our example we fix $ESS_{0}=N/2$, this is an arbitrary common rule of thumb.}, resempling applies
\begin{itemize}
\item Draw a sample of size N, $x_{t}^{(1)},...,x_{t}^{(N)}$,from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{itemize}
```{r}
PFfun<-function(data,N,m0,C0,tau,sigma,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

@@

    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x<-x[index]
      w<-rep(1/N,N)
    }else{}

@@

```{r}
GPFfun<-function(data,N,m0,C0,tau,sigma,r){
 if(missing(r)){r=2}else{}
   xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

@@

```{r}
APFfun<-function(data,N,m0,C0,tau,sigma,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

@@

    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x1<-x1[index]
      w<-rep(1/N,N)
    }else{}

@@

\end{itemize}
```{r}
LWfun<-function(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r){
  if(missing(r)){r=2}else{}
  xs     = rnorm(N,m0,sqrt(C0))
  if(unif==T){
  pars   = cbind(runif(N,0,10),runif(N,0,10))}else{}
  pars   = cbind(rgamma(N,shape=alphav,scale=betav),rgamma(N,shape=alphaw,scale=betaw))
  a      = (3*delta-1)/(2*delta)
  h2     = 1-a^2
  parss  = array(0,c(N,2,n))
  xss    = NULL
  ws     = NULL
  ess    = NULL
  w      = rep(1/N,N)
  for (t in 1:length(data)){
    meanV = weighted.mean(pars[,1],w)
    varV  = weighted.mean((pars[,1]-meanV)^2,w)
    meanW = weighted.mean(pars[,2],w)
    varW  = weighted.mean((pars[,2]-meanW)^2,w)

@@

    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      xs<-xs[index]
      pars<-pars[index,]
      w<-rep(1/N,N)
    }else{
      xs<-xs
      pars<-pars
    }

@@

# COMPARISON (ANOTHER SECTION)
```{r}
library(ggpubr)
library(kableExtra)

Kalmanvarplot<-ggplot(DLM.df,aes(x=timeframe))+
  geom_line(aes(y=C))+
  labs(x="Time",
       y="")+
  ggtitle("Kalman Filter")+
  theme_bw()

Varplot<-function(dataframe,fun,title){

ggplot(dataframe,aes(x=timeframe))+
geom_line(aes(y=sd))+
  labs(x="Time",
       y="")+
  ggtitle(title)+
  theme_bw()

Varsisplot<-Varplot(df,sis,"SIS")
Varpfplot<-Varplot(df,pf,"PF")
Varapfplot<-Varplot(df,pf,"APF")
Varlwplot<-Varplot(df,pf,"LW")

ggarrange(Kalmanvarplot,Varsisplot,Varpfplot,Varapfplot,Varlwplot)

```{r}
truex<-x
estimatedx<-matrix(NA,ncol=6,nrow=6)
colnames(estimatedx)<-c("N","Threshold","KF","PF","APF","LWF")
estimatedx[,1]<-c(20,100,500,1000,1000,1000)
estimatedx[,2]<-c(0.5,0.5,0.5,0.5,0.25,0.1)
RMSE<-function(x,xhat){sqrt(mean((x - xhat)^2))}

i=1
for(N in c(20,100,500)){

DLM1<-DLM(y,sigma2,tau2,m0,C0)
pf1<-PFfun(y,N,m0,C0,tau,sigma)
apf1<-APFfun(y,N,m0,C0,tau,sigma)
lwf1<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T)

estimatedx[i,3]<-RMSE(truex,DLM1$m)
estimatedx[i,4]<-RMSE(truex,Filtervalues(pf1)$mean)
estimatedx[i,5]<-RMSE(truex,Filtervalues(apf1)$mean)
estimatedx[i,6]<-RMSE(truex,Filtervalues(lwf1)$mean)
i=i+1

i=4
for(k in c(2,5,10)){
  N=1000
DLM2<-DLM(y,sigma2,tau2,m0,C0)
pf2<-PFfun(y,N,m0,C0,tau,sigma,r=k)
apf2<-APFfun(y,N,m0,C0,tau,sigma,r=k)
lwf2<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T,r=k)

estimatedx[i,3]<-RMSE(truex,DLM2$m)
estimatedx[i,4]<-RMSE(truex,Filtervalues(pf2)$mean)
estimatedx[i,5]<-RMSE(truex,Filtervalues(apf2)$mean)
estimatedx[i,6]<-RMSE(truex,Filtervalues(lwf2)$mean)
i=i+1

```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
kbl(estimatedx[1:3,], align="c", longtable = T, booktabs = T, caption="RMSE",digits=3)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
kbl(estimatedx[4:6,], align="c", longtable = T, booktabs = T, caption="RMSE",digits=3)
```

# STOCHASTIC VOLATILITY

Basic specification of Stochastic Volatilty Model
\begin{align}
y_{t}|x_{t} & \sim N(0,e^{x_{t}}) \\
x_{t}|x_{t-1} & \sim N(\alpha+\beta x_{t-1},\tau^2)
\end{align}
\
```{r include=FALSE}
#############################
#
#   y(t) = exp(h(t)/2)*v(t)
#   x(t) = alpha + beta*x(t-1) + tau*w(t)
#
#   v(t) ~ N(0,1)
#   w(t) ~ N(0,1)
#   x(0) ~ N(0,100)
#
#   PRIOR ON alpha,beta,tau2

# Simulating the data
# -------------------
set.seed(12345)
n     =  50
alpha = -0.0031
beta  =  0.9951
tau2  =  0.0074
tau   = sqrt(tau2)
y     = rep(0,n)
x     = rep(0,n)
x[1]  = alpha/(1-beta)
y[1]  = rnorm(1,0,exp(x[1]/2))
for (t in 2:n){
  x[t] = alpha + beta*x[t-1] + rnorm(1,0,tau)
  y[t] = rnorm(1,0,exp(x[t]/2))
}
alpha.true = alpha
beta.true  = beta
tau2.true  = tau2

timeframe<-c(1:length(y))
dfsv<-data.frame(timeframe,y,x)
```

```{r echo=FALSE}
n=50
tau2=1
sigma2=1
sigma=sqrt(sigma2)
tau=sqrt(tau2)
x0=0

ggplot(dfsv,aes(x=timeframe))+
  geom_line(aes(y=y, linetype="Observations"))+
  geom_line(aes(y=x, linetype="True States"))+
  scale_linetype_manual(name="",
                      values=c("Observations" = 1,
                               "True States" = 3))+
  labs(x="Time",
       y="")+
  ggtitle("States and Observations")+
  theme_bw()+
  theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
```

\textbf{Particle Filter}
\
```{r}
SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

    x<-rnorm(N,alpha+beta*x,tau)
    w1<-w*dnorm(data[t],0,exp(x/2))

    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x<-x[index]
      w<-rep(1/N,N)
    }else{}

\textbf{Auxiliary Particle Filter}

```{r}
SVAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

    weight = w*dnorm(data[t],0,exp(x/2))
    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
    x1   = rnorm(N,alpha+beta*x[k],tau)
    lw  = dnorm(data[t],0,exp(x/2),log=TRUE)-dnorm(data[t],0,exp(x/2),log=TRUE)
    w   = exp(lw)
    w   = w/sum(w)
    ESS  = 1/sum(w^2)

    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x1<-x1[index]
      w<-rep(1/N,N)
    }else{}

\textbf{Liu and West}

```{r include=FALSE}
# Data and prior hyperparameters 
# ------------------------------
m0      = 0.0
C0      = 0.1
sC0     = sqrt(C0)
ealpha  = alpha
valpha  = 0.01
ebeta    = beta
vbeta    = 0.01
nu      = 3
lambda  = tau2

# Liu and West filter
# -------------------
set.seed(8642)
N      = 1000
```

```{r}
SVLWfun<-function(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda){

xs     = rnorm(N,m0,sqrt(C0))
pars   = cbind(rnorm(N,ealpha,sqrt(valpha)),rnorm(N,ebeta,sqrt(vbeta)),
               log(1/rgamma(N,nu/2,nu*lambda/2)))
delta  = 0.75
a      = (3*delta-1)/(2*delta)
h2     = 1-a^2
parss  = array(0,c(N,3,n))
xss    = NULL
ws     = NULL
ESS    = NULL
w      = rep(1/N,N)
for (t in 1:length(data)){
  mpar        = apply(pars,2,mean)
  vpar        = var(pars)
  ms          = a*pars+(1-a)*matrix(mpar,N,3,byrow=T)
  mus         = pars[,1]+pars[,2]*xs 
  weight      = w*dnorm(data[t],0,exp(mus/2))
  k           = sample(1:N,size=N,replace=T,prob=weight)
  ms1         = ms[k,] + matrix(rnorm(3*N),N,3)%*%chol(h2*vpar)
  xt          = rnorm(N,ms1[,1]+ms1[,2]*xs[k],exp(ms1[,3]/2))
  w           = dnorm(y[t],0,exp(xt/2))/dnorm(y[t],0,exp(mus/2))
  w           = w/sum(w)
  ESS         = 1/sum(w^2)

  if(ESS<(N/2)){
    index<-sample(N,size=N,replace=T,prob=w)
    xs<-xt[index]
    pars<-ms1[index,]
    w<-rep(1/N,N)
  }else{
    xs<-xt
    pars<-ms1
  }

  xss         = rbind(xss,xs)
  parss[,,t]  = pars 
  ws          = rbind(ws,w)
}
return(list(xs=xss,pars=parss,ws=ws))
}
```

```{r echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
#Kalman Filter
#-------------
m0=0
C0=100
filtval<-DLM(y,sigma2,tau2,m0,C0)
m<-filtval$m
C<-filtval$C
Low = m + qnorm(0.025)*sqrt(C)
Up = m + qnorm(0.975)*sqrt(C)

```{r include=FALSE}
N=1000
svpf<-SVPFfun(y,N,m0,C0,alpha,beta,tau)
svapf<-SVAPFfun(y,N,m0,C0,alpha,beta,tau)
#svlw<-SVLWfun(y,N=5000000,m0,C0,ealpha=0,valpha=1,ebeta=1,vbeta=1,nu=1,lambda=1)
```

```{r}
#Filtplot(dfsv,,"")
```
